EBASE manuscript

### 2023-04-18 manu-draft.qmd
added some initial text to discussion

As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) than using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative). Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

@@

-   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
-   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.

>>>>

As noted in the previous paragraph, metabolic estimates between the methods were generally more similar based on RMSE results using the detided dissolved oxygen time series ([@fig-apacmpfig]b) than using the observed dissolved oxygen as input ([@fig-apacmpfig]a). This suggests that tidal effects on continuous monitoring data can severely violate assumptions required for methods of estimating ecosystem metabolism [as described in detail in @Beck15]. These effects are easily identified with estimates from the Odum method as negative values for P and positive values for R, i.e., "anomalous" values. The effects of tidal advection on estimates from EBASE were more subtle as anomalous values cannot be returned based on constraints of the priors (i.e., none can be negative), although the range of values for the metabolic estimates was much larger with observed dissolved oxygen. Overall, the results suggest that detiding the dissolved oxygen time series prior to estimating metabolic rates produces more stable estimates with smaller ranges (`r run_reference(id = "tbl-apacmptab")`), regardless of the method used.

@@

EBASE provides several computational advantages over BASE implemented in the R package that was created for the purposes herein.  Most importantly, EBASE was developed to allow different model optimization lengths, whereas the BASE method was developed to estimate results for only a single day.  Allowing EBASE to incorporate more than a day of observations in the optimization can likely produce more accurate and precise estimates for the metabolic parameters.  Physical and biological characteristics of an ecosystem that affect the metabolic rates are likely acting at time scales longer than a day, e.g., wind patterns that can affect gas exchange as manifested in the $b$ parameter may persist for several days.  Exposing the model to additional observations may produce more stable results as the parameters for the best fit are estimated by the Bayesian routine as a function of the data that are evaluated each optimization period.  However, we cannot provide a precise recommendation on the number of days to use for the optimization period as the exact length of time that ecosystem characteristics can affect metabolic rates may not be known and likely varies by location.  Our example using the synthetic time series from Apalachicola Bay suggests that results were similar for the 7 and 30 day optimization period. As such, using one week of data for each optimization period is likely appropriate for this location.  Using one day is not recommended given the issues stated above (i.e., unstable results), whereas an optimization period longer than a month may miss important events in the time series that can affect the metabolic estimates.

The EBASE R package can also accommodate missing observations in the input data. This allows the estimation of metabolic rates for time series that cover relatively long periods of time (e.g., more than year), when data gaps are likely to occur for several reasons (e.g., equipment malfunction, routine maintenance, etc.).  The R package accommodates data gaps using linear interpolation for any required parameter used by EBASE.  This allows the Bayesian routine to provide continuous estimates across the time series.  The results are then post-processed such that any metabolic estimates that exceed a threshold for interpolated observations are removed from the output.  The analyst can determine the appropriate length of time to use for excluding results based on interpolated values.  The default setting for EBASE is 12 hours, where results for an entire optimization period are removed if any interpolated period exceeded 12 hours.  Although this does not apply to the synthetic time series, some of the estimates using the 2012 observed data for Apalachicola Bay were excluded from the analysis.  This feature allows EBASE to be used more readily because an analyst does not need to manually screen data that include missing values.

-   Primary modifications (Production is linearly related to PAR (aPAR), not saturation point Respiration is constant (not temp dependent), Uses a gas exchange commonly used in oceanic systems (not littoral - streams/lakes as in Grace))
-   Describe how EBASE deals with gaps and flexibility of the model to accommodate diff optimization periods
-   b is sketchy, how to improve? Either fix or develop JAGS model further to include additional data, e.g. tidal height.
-   detiding
-   other main conclusions (from RGN comments on draft?)
